Food-web structure, stochasticity and response diversity drive temporal community stability in complex communities

### 2024-04-04 paper/manuscript.Rmd
Update manuscript and supplementary figures

---
title: Food-web structure, stochasticity and response diversity drive temporal community stability in complex communities
author: "Alain Danet, (Sonia Kéfi, Thomas F Johnson), Andrew P. Beckerman"
date: ""
output:
  officedown::rdocx_document:
    number_sections: false
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    spacing: double
    number_sections: false
always_allow_html: true
bibliography: references.bib
link-citations: true
csl: ecology-letters.csl
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
---

@@

@@

Current biotic homogenization and increasing frequency of extreme events are
projected to dramatically destabilize ecological communities. Understanding how
complex communities such as food-webs respond to such threats is crucial to take
adequate conservation measures. We then investigated how response diversity,
environmental stochasticity, and food-web structure drive temporal stability of
biomass in complex food-webs, using a stochastic bioenergetic food-web model. We
found that higher average trophic level, omnivory degree and generalism
increased  average population stability while species richness had little
effects. At the opposite higher average trophic level, omnivory degree and
generalism decreased asynchrony and compensatory dynamics raising trophic
interactions. We further found that that high interaction strength drive
negative stability-species richness when response diversity is low but that high
response diversity always drive positive stability-richness relationship. Our
study reveals that response diversity and food-web structure drive temporal
stability and further improve our understanding of biodiversity-stability
relationships in complex trophic communities.

@@

Global environmental change constitutes a suite of major threats to biodiversity
and ecosystem functioning [@lee_ipcc_2023; @simmons_refocusing_2021]. A central
feature of these threats is that they are predicted to dramatically increase
variability in the environment that species face [@lee_ipcc_2023] with potential
impacts on the stability of communities and associated impacts on biodiversity
and functioning in these communities. Assessing the impacts of this variability
on communities is challenging for several reasons.

First, the stability of communities is often evaluated using temporal
variability the metric of interest (e.g. abundance or biomass), because most
natural communities face persistent perturbations so that other stability
metrics such as resilience and resistance are not always applicable
[@arnoldi_resilience_2016]. Understanding temporal community variability,
measured as the inverse of the coefficient of variation
[@tilman_biodiversity_2006; @roscher_identifying_2011;
@thibaut_understanding_2013; @zhao_biodiversity_2022], requires scaling down
the focus to populations. Community stability can be mathematically partitioned
into average population stability and asynchrony [@loreau_species_2008;
@thibaut_understanding_2013; @zhao_biodiversity_2022], asynchrony measuring how
much each species fluctuations are compensating each other at the community
level. Population stability and asynchrony depend both on the number of species
and their dominance [@loreau_species_2008; @thibaut_understanding_2013;
@zhao_biodiversity_2022], the interactions among them [@thebault_trophic_2005;
@thebault_relationship_2006; @loreau_species_2008; @danet_species_2021;
@eschenbrenner_diversity_2023] and on the way that species respond to
environmental perturbations [response diversity, @ives_stability_2000;
@elmqvist_response_2003; @ross_how_2023]. Therefore, accounting for the effects
on community structure on population stability and asynchrony is crucial to
improve our understanding of community stability in face of environmental
change.

Average population stability is decreased by external forcing, such as the
intensity and the frequency of perturbations [@loreau_species_2008]. Species richness can
decrease [@tilman_biodiversity_2006; @loreau_species_2008], increase
[@houlahan_negative_2018], or have little effects on population stability. This
effect of species richness on population stability depends on the evenness of
the community such as dominant species contribute more the average of the
community than subordinate species. At the core of ecological theory,
population stability depends on internal forcing, here species interactions. The
topological structure and the distribution of interaction strength have tremendous
effects on population stability, such as strong interaction strength are
generally destabilising [@mccann_weak_1998; @mccann_reevaluating_1997;
@thebault_trophic_2005]. In particular for food-webs, top-down and bottom-up
controls can increase population stability.

Asynchrony in turn can be partitioned into portfolio effects and compensatory
dynamics and is mostly determined by internal factors. Portfolio effects stems
from the statistical averaging of independent population fluctuations
[@doak_statistical_1998; @zhao_biodiversity_2022], thereby increasing
asynchrony. Compensatory dynamics in turn captures population fluctuations that
are more synchronous or asynchronous than expected for independent fluctuations,
and can stem from species interactions and species responses to perturbations
[@zhao_biodiversity_2022]. Response diversity captures how among species’
differences in their response to the environmental variation, can underpin
asynchronous dynamics if their responses are different but synchronous if their
responses are similar. Theory and empirical data emerging around response
diversity suggests it is  “core ecological mechanism underpinning many of the
statistical explanations of the diversity–stability relationships”
[@ives_stability_2000; @elmqvist_response_2003; @ross_how_2023]. In turn,
the distribution and strength of species interactions can also modulate
asynchrony through compensatory dynamics such as a strong asymmetry of consumer
preference can generate asynchrony in resource fluctuations [@mccann_weak_1998;
@vasseur_environmental_2007], and a generalist consumer can in turn generate
synchrony among its prey [@raimondo_interspecific_2004]. Competition can also
have synchronising effects in strong species growth rates regime
[@loreau_species_2008]. Finally, portfolio and compensatory effects can also be
affected by the population evenness such as fluctuations of a dominant
species will hardly be compensated by a subordinate species. Improving our
understanding of community stability facing perturbations therefore requires to
articulate the effect response diversity, species interactions, species
richness and evenness on the different components of stability.

The stability of ecological communities facing perturbations stems from two
parallel bodies of work: species interactions and response diversity. These two
bodies of work share the common thread of community structure acting through a
filter to drive stability. However, the filter is different in each space; in
the first, the filter is the topology and the strength of species interactions
and in the second variation in species’ responses. Moreover, the theory has been
developed largely in theoretical or empirical communities characterised by a
limited array of interactions (e.g. often lacking in consumers) and by limited
species diversity/richness. Thus, ecology lacks a body of theory integrating
across structure and response diversity frameworks and reflecting species rich
communities. Here, we integrate these two perspectives on diversity - stability
to evaluate the combined effects of community structure and response diversity on
temporal stability in species rich, complex communities characterised by
competition and predation.

The ability to do this is supported by a key statistical advance: it is now
possible to partition asynchrony into two component sources, the statistical
averaging effect (SAE) and compensatory dynamics effect (CPE)
[@zhao_biodiversity_2022]. SAE quantifies the portfolio effect emerging from
independent species fluctuations, while CPE describes compensatory dynamics
emerging from species interactions and differential responses to environmental
stochasticity. This allows an unprecedented opportunity to obtain mechanistic
insight about the effects of stressors and food-web structure on stability in
complex, species rich communities.

Our primary objective is to derive theory that predicts how the components of
community stability, that can differentially drive patterns of stability, deliver novel
insights into the source and patterns of diversity-stability relationships in
complex communities facing environmental stochasticity. Our approach delivers
revised insight into diversity-productivity relationships in species rich
communities when both community structure per se, and species
responses can vary. In the following sections, we begin by introducing the
multi-species model and the features of the model that allow us to integrate
across variation in community structure and response diversity. We
then introduce assumptions we used to implement environmental stochasticity,
response diversity, variation in community structure and variation in the
distribution of interaction types and strengths. Next we define our in-silico
experimental design that specifically aligns with three core questions and
associated analyses: (1) What are the direct effects of food-web structure and
interaction strength on population stability, compensatory dynamics,
and statistical averaging effects (i.e. portfolio effects)? (2) What are the
total effects of food-web structure, response diversity, environmental
stochasticity on community stability. Finally, merging the two parallel bodies of
work, we asked (3) How do response diversity and food-web structure interact to
drive the sign of the relation between community stability and
species richness?

@@

We simulated the dynamics of complex food-webs using the allometric bioenergetic
model [@yodzis_body_1992; @brose_allometric_2006; @delmas_simulations_2017].
This model is widely used to simulate the dynamics of complex ecological
communities because of the simplicity of its parametrisation and that its
parameters are rooted in the metabolic theory of ecology [@brown_toward_2004].
The model describes species biomass dynamics over time with two master equations (eq. \@ref(eq:dbdt)), one for the
primary producers (eq. \@ref(eq:prod)) and one for the consumers (eq. \@ref(eq:prod)):

@@

The consumers gain biomass by consuming resources (i.e. their preys) at a rate
that depends on their mass specific metabolic rate ($x_i$),  maximum consumption
rate ($y_i$) and on the functional response ($F_ij$) describing how the rate of
consumption of a consumer $i$ on a resource $j$ vary with the biomass of this resource.
Consumers then loses biomass by being consumed, $e_ij$ being the assimilation
efficiency of the consumer $j$ on the resource $i$. Consumers also lose biomass
over time through metabolic losses ($-x_iB_i$) and by natural mortality due to
environment ($-d_iB_i$). The primary producers gain biomass over time with a
growth rate ($r_i$, eq. \@ref(eq:prod)) and a functional response ($G_i$)
describing how the growth of the producers vary with their biomass. They lose
biomass by being consumed by consumers and by natural mortality.

@@

The growth function of the primary producers (eq. \@ref(eq:growth)) has the shape of a logistic growth
where the growth rate is maximal and null when
$\sum_{j \in \{\mathrm{prod}\}} \alpha_{ij} B_j$ are respectively close to 0 and 1. $\alpha_{ij}$ is the
per capita effect of the producer $j$ on the producer $i$, $K_i$ being the
carrying capacity for the producer $j$. The functional response of a consumer
feeding on a resource (eq. \@ref(eq:holling)) depends on the relative preference
of the consumer on the resource ($\omega_ij$) is limited by a half-saturation
rate ($B_0$), intraspecific interference coefficient ($c_i$), and on the
availability of its other resources
($\sum_{k \in \{\mathrm{res}\}_i} \omega_{ik} B_k^h$). Finally, the $h$ exponent
controls the shape of the functional response from a saturating function
($h = 2$, Holling type II) to a sigmoid ($h = 3$, Holling type III).

@@

We introduced environmental stochasticity by
considering stochastic variation of mortality rates. Doing so allows to consider
stochasticity without modifying species interactions themselves
[@vasseur_environmental_2007]. The stochastic part of the mortality rates
($\epsilon_{i, t}$) follow a normal distribution 0 centered
($\epsilon_{i, t} \sim N(0,\sigma^2_e)$) with a variance ($\sigma^2_e$). The mortality rates each
species at the time $t$ equals to $d_{i,t} = d_i e^{\epsilon_{i, t}}$. We did so
to ensure to not have negative mortality rates [@vasseur_environmental_2007]. We
simulated $\epsilon$ using an Ornstein-Uhlenbeck process, which is a modified
Brownian motion where the stochastic values tend to come back to the central
value (i.e. $\epsilon = 0$) such as $\epsilon_t$ varies according to this
stochastic differential equation: $d\epsilon_t = (0 - \epsilon_t)d_t + \sigma_e
dW_t$, where $dW_t$ is a Brownian motion. The strength of environmental
stochasticity was then controlled by $\sigma_e$.

@@

\begin{equation}
\begin{aligned}
\epsilon_{i, t} =
  \begin{bmatrix}
     \rho_{11}\sigma^2_e & \rho_{12}\sigma^2_e & \cdots & \rho_{1n}\sigma^2_e \\
     \rho_{21}\sigma^2_e & \rho_{22}\sigma^2_e & \cdots & \rho_{2n}\sigma^2_e \\
     \vdots  & \vdots  & \ddots & \vdots  \\
     \rho_{n1}\sigma^2_e & \rho_{n2}\sigma^2_e & \cdots & \rho_{nn}\sigma^2_e
  \end{bmatrix}
  \begin{bmatrix}
     a_{1,t} \\
     a_{2,t} \\
     \vdots  \\
     a_{n,t}
  \end{bmatrix}
S_{com} = S_{pop} \times \phi
(\#eq:rho)
\end{aligned}
\end{equation}

@@

}
pS <- get_sim_par_val(x = "S")
pct <- round(get_sim_par_val(x = "ct"), 2)
pZ <- get_sim_par_val(x = "Z")
ph <- get_sim_par_val(x = "h")
pe <- get_sim_par_val(x = "env_stoch")
prd <- get_sim_par_val(x = "resp_div")
pr <- get_sim_par_val(x = "rho")
```
\pagebreak

@@

To vary interaction strength and mortality rates (i.e. allometrically
scaled mortality rate $d$) in the community, we varied the Predator-Prey
Mass Ratio (PPMR) from $`r pZ["min"]`$ to $`r pZ["max"]`$ (Fig. \@ref(fig:met)b). We further varied
the functional response of consumer from type II to type III (i.e. from
$h = `r ph["min"]`$ to $h = `r ph["max"]`$, eq. \@ref(eq:holling)) as it was shown to
have a tremendous role on stability [@brose_allometric_2006]. We set a global
carrying capacity of $K' = 10$ to ensure that consumers were not limited by the
biomass input from primary producers. We standardised the carrying capacity by
the number of primary producers and by the interspecific competition among
producers to ensure that the effects of species richness on temporal stability
are not driven by the increase in the number of primary producers, thereby
trivially increase biomass input for consumers. We then standardised $K'$ such as
$K_i = \frac{1 + \alpha_{ij} (S - 1)}{S}K'$ [@ives_stability_2000].

We numerically solved the stochastic differential equation systems during 5000
timesteps and collected the last 500 timesteps. We set $d_t = .1$, and $d_t =
.05$ when instability was detected during the simulations due to stiff changes
in biomass. We set extinction threshold to $10^{-6}$ and if an extinction
happened in the last 500 timesteps, we continued to run the simulation for
another 1000 timesteps until there was no extinction events. When we found
disconnected species at the end of the simulation (i.e. a primary producer
without consumer or a consumer without prey), we removed them and re-run the
simulation for another 1000 timesteps.

@@

We measured food-web structure and dynamics that have been shown to be linked
with stability. We collected total biomass, species richness, connectance,
average trophic level weighted by biomass, average omnivory and average per
capita interaction strength. Connectance was computed as $C = L / S(S-1)$, $L$
being the number of trophic links and $S$ being the number of species in the
community. Species trophic level was computed recursively from the bottom of the
food-web, such as the trophic level of a consumer was equal to the average
trophic level of its resources added to 1, the trophic level of primary
producers being set to 1 [@christensen_ecopath_1992]. Average trophic level was then equal to:
$1/\hat{B}_i\sum_{i}\mathrm{TL}_i$, $\mathrm{TL}_i$ being the trophic level and
$\hat{B}_i$ the average biomass of the species $i$. The degree of omnivory of a
consumer was computed  such as the sum of squares of its
resource trophic levels weighted by the relative preference of the consumer for
each resource: $\sum_i (\mathrm{TL}_i - \hat{\mathrm{TL}})^2 \omega_i/ \sum \omega_i$ [@christensen_ecopath_1992].
Interaction strength was computed was quantified as the biomass fluxes going
from a resource to a consumer, such as: $I = x_i y_i B_k F_{ij}$ (eq.
\@ref(eq:cons)), and averaged over time. Per capita interaction strength was
obtained by dividing average interaction strength by the average biomass of the
consumer. Our way of measuring per interaction strength is very similar to
previous metrics used in previous studies, at the difference here that we
average fluxes over time and not the maximal interaction strength
[@mccann_weak_1998]. Finally, the average interaction strength of a community
was computed as the average per capita interaction strength at the exclusion of
null interaction strengths (i.e. excluding absence of trophic interactions).

@@

To further get a mechanistic insight about the effects of stressors and food-web
structure on stability, we partitioned asynchrony into statistical averaging
effect (SAE) and compensatory dynamics effect (CPE) [@zhao_biodiversity_2022].
SAE quantities the portfolio effect emerging from independent species
fluctuations, while CPE describes compensatory dynamics emerging from species
interactions and differential responses to environmental stochasticity.

@@

Using a structural equation model, we disentangled the complex effects of
food-web structure, environmental stochasticity and response diversity on the
different facets of temporal stability of biomass (i.e. population stability,
statistical averaging effects and compensatory dynamic effects). We set the
paths of the path of the model according to theoretical expectations. We tested
the effects of initial species richness, initial connectance, Predator-Prey size
ratio on secondary structure of the food-webs, i.e. realized species richness
and connectance, the average trophic level, the average omnivory and the
interaction strength. Further, we expected those food-web structure to alter
population stability and synchrony. Grounded in theoretical ecology, we expected
that higher average trophic level, average omnivory, lower interaction strength,
and type III functional response (relative to type II) increase population
stability [@mccann_weak_1998; @mccann_reevaluating_1997;
@shanafelt_stability_2018]. Higher Predator-Prey size ratio was also expected to
increase population stability [@brose_allometric_2006]. Finally, the effect of
environmental stochasticity was assumed to be mainly associated to population
stability (i.e. by increasing amplitude of population fluctuations).

Regarding asynchrony, we tested the effects of species richness, connectance,
average trophic level and interaction strength on statistical averaging effects.
We expected that those same food-web structural properties affected
compensatory dynamics due to interactions ($CPE_{int}$). Finally, we tested the
effects of species richness and species response diversity in mortality rates on
the compensatory dynamics due to stochastic fluctuations ($CPE_{env}$). We
completed the structural equation model with the mathematical relationship
linking compensatory dynamics and statistical averaging effects to asynchrony,
and from asynchrony and population stability to community stability.

>>>>

---
title: Food-web structure, stochasticity and response diversity drive temporal community stability in complex communities
author: "Alain Danet (1), Sonia Kéfi (2,3), Thomas F Johnson (1), Andrew P. Beckerman (1)"
date: ""
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    spacing: double
    number_sections: false
always_allow_html: true
bibliography: references.bib
link-citations: true
csl: ecology-letters.csl
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
---

@@

- 1 School of Biosciences, Ecology and Evolutionary Biology, University of Sheffield, Western Bank, Sheffield, S10 2TN, UK
- 2 ISEM, CNRS, Univ. Montpellier, IRD, EPHE, Montpellier, France
- 3 Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, NM 87501, USA

@@

Empirical and theoretical evidence suggests that current and projected
environmental changes and biotic homogenisation will dramatically destabilise
ecological communities. However, this evidence is largely based on
oversimplified communities (i.e. food chains, guild assemblages), which lack the
complexity of natural communities.We characterise the dynamics of complex
food-webs under a stochastic bioenergetic food-web model, and investigate how
response diversity, environmental stochasticity, and food-web structure drive
temporal stability. We found that higher average trophic level, omnivory degree
and generalism increased average population stability, but decreased asynchrony,
triggering compensatory dynamics that raised trophic interactions.  Species
richness had little effect on average population stability, but a large one on
synchrony. Further, when response diversity is low, high interaction strength,
average trophic level, and connectance all lead to negative stability-species
richness relationships, but when response diversity is high, we observe positive
stability-species richness relationships regardless of the food-web structure.
Our study reveals that response diversity and food-web structure drive temporal
stability and further improve our understanding of biodiversity-stability
relationships in complex trophic communities.

@@

Global environmental change constitutes a suite of major threats to biodiversity
and ecosystem functioning [@lee_ipcc_2023; @simmons_refocusing_2021]. A central
feature of these threats is that they are predicted to dramatically increase
variability in the environment that species face [@lee_ipcc_2023] with potential
impacts on the stability of communities and associated impacts on biodiversity
and functioning in these communities. Assessing the impacts of this increased
variability on ecological communities is challenging for several reasons.

First, the stability of communities is often evaluated using the temporal
variability of a metric of interest (e.g. abundance or biomass), because other
stability metrics such as resilience and resistance are not always applicable,
as most natural communities face persistent perturbations [@arnoldi_resilience_2016].
Understanding this temporal community variability (hereafter referred to as
community stability), measured as the inverse of the coefficient of variation
[@tilman_biodiversity_2006; @roscher_identifying_2011;
@thibaut_understanding_2013; @zhao_biodiversity_2022], requires scaling down the
focus from communities to populations.
Community stability can be mathematically partitioned into average population
stability and asynchrony [@loreau_species_2008;
@thibaut_understanding_2013; @zhao_biodiversity_2022] - asynchrony measuring how
much each species fluctuations’ are compensating for each other at the community
level. Population stability and asynchrony both depend on the number of species
and their dominance [@loreau_species_2008; @thibaut_understanding_2013;
@zhao_biodiversity_2022], on the interactions among species [@thebault_trophic_2005;
@thebault_relationship_2006; @loreau_species_2008; @danet_species_2021;
@eschenbrenner_diversity_2023] and on the way species respond to environmental
perturbations [@ives_stability_2000; so-called response diversity,
@elmqvist_response_2003; @ross_how_2023]. Therefore, accounting for the effects
of community structure on population stability and asynchrony is crucial to
improve our understanding of community stability in facing environmental change.

Average population stability decreases with environmental stochasticity, such as the
intensity and the frequency of perturbations [@loreau_species_2008]. Theorerical
and empiral evidence suggests that species richness can decrease
[@tilman_biodiversity_2006; @loreau_species_2008], increase
[@houlahan_negative_2018], or have little effects on population stability. This
effect of species richness on population stability depends on the evenness of
the community since dominant species contribute more to the average of the
community than subordinate species. Species interactions also have tremendous
effects on population stability, such as strong interaction strength are
generally destabilising [@mccann_weak_1998; @mccann_reevaluating_1997;
@thebault_trophic_2005]. The shape of food-webs can also modulate population
stability, such as higher trophic level was generally found to be stabilising
[@shanafelt_stability_2018; @danet_species_2021; @eschenbrenner_diversity_2023],
as top-down and bottom-up controls can increase population stability
[@shanafelt_stability_2018].

Asynchrony is mostly determined by the characteristics of the network. Response
diversity captures how species’ differences in their response to the
environmental variation can lead to asynchronous dynamics if their responses are
different but synchronous dynamics if their responses are similar. Theory and
empirical data emerging around response diversity suggest that it is a “core
ecological mechanism underpinning many of the statistical explanations of the
diversity–stability relationships” [@ives_stability_2000;
@elmqvist_response_2003; @ross_how_2023]. In turn, the network structural
properties and strength of species interactions can also modulate asynchrony
through compensatory dynamics such as a strong asymmetry of consumer preference
can generate asynchrony in resource fluctuations [@mccann_weak_1998;
@vasseur_environmental_2007], and a generalist consumer can in turn generate
synchrony among its prey [@raimondo_interspecific_2004]. Competition can also
have synchronising effects in strong species growth rates regime
[@loreau_species_2008]. Finally, portfolio and compensatory effects can also be
affected by the population evenness such as fluctuations by dominant species
will hardly be compensated by a subordinate species
[@thibaut_understanding_2013; @zhao_biodiversity_2022]. Improving our
understanding of community stability facing perturbations therefore requires
articulating the effects of response diversity, species interactions, species
richness and evenness on the different components of stability. Predicting the
effects of species richness on community stability thus requires understanding
the relationship between species richness, population stability and asynchrony.

Empirical evidence suggest by increasing asynchrony, species richness
drives positive relationships between community stability and species richness
[@tilman_biodiversity_2006; @isbell_biodiversity_2009]. In contrast population stability was
found to either decrease, increase, or not change with species richness
[@houlahan_negative_2018; @tilman_biodiversity_2006; @roscher_identifying_2011].
We note, however, that evidence for both asynchrony-richness and population
stability-richness is derived from data on species assemblages.  These data thus
do not account for food-web structure, which can have significant impact on
population stability and asynchrony [but see @danet_species_2021]. Thus, ecology
lacks a body of theory of temporal stability integrating across structure and
response diversity frameworks and reflecting species rich communities. Here, we
integrate these two perspectives on diversity- stability to evaluate the
combined effects of community structure and response diversity on temporal
stability in species rich, complex communities characterised by competition and
predation.

The ability to do this is supported by a key statistical advance: it is now
possible to partition asynchrony into two component sources: the portfolio
effects (rising from statistical averaging effects, SAE) and the compensatory
dynamics effect (CPE) [@zhao_biodiversity_2022]. Portfolio effects stem from the
statistical averaging of independent population fluctuations [@doak_statistical_1998;
@zhao_biodiversity_2022], thereby increasing asynchrony. Compensatory dynamics
in turn capture population fluctuations that are more synchronous or
asynchronous than expected for independent fluctuations, and can stem from
species interactions and species responses to perturbations [@zhao_biodiversity_2022].
This allows an unprecedented opportunity to obtain mechanistic insight about the
effects of stressors and food-web structure on stability in complex, species
rich communities.

Our primary objective is to derive theory that predicts how the components of
community stability, that can differentially drive patterns of stability,
deliver novel insights into the source and patterns of diversity-stability
relationships in complex communities facing environmental stochasticity, a
necessary ingredient to generate species fluctuations. Our approach delivers
revised insight into diversity-stability relationships in species rich
communities when both community structure per se, and species responses can
vary. In the following sections, we begin by introducing the multi-species model
and the features of the model that allow us to integrate across variation in
community structure and response diversity. We then introduce assumptions we
used to implement environmental stochasticity, response diversity, variation in
community structure and variation in the distribution of interaction types and
strengths. Next, we define our in-silico experimental design that specifically
aligns with three core questions and associated analyses. Considering four
drivers, namely food-web structure, interaction strength, environmental
stochasticity and response diversity, we investigated (1) What are their direct
effects on population stability, portfolio and compensatory effects ? (2) What
are their total effects on community stability, population stability and
asynchrony? Finally, merging the two parallel bodies of work, we asked (3) How
do they interact to drive the sign of the relation between community stability
and species richness?

@@

We simulated the dynamics of complex food-webs using the allometric bioenergetic
model [@yodzis_body_1992; @brose_allometric_2006; @delmas_simulations_2017]. This model
is widely used to simulate the dynamics of complex ecological communities
because of the simplicity  and because it can be parametrized using  the
metabolic theory of ecology [@brown_toward_2004]. The model describes species
biomass dynamics over time with two master equations (eq. \@ref(eq:dbdt)), one
for the primary producers (eq. \@ref(eq:prod)) and one for the consumers (eq.
\@ref(eq:prod)):

@@

The consumers gain biomass by consuming resources (i.e. their preys) at a rate
that depends on their mass specific metabolic rate ($x_i$),  maximum consumption
rate ($y_i$) and on the functional response ($F_ij$) describing how the rate of
consumption of a consumer $i$ on a resource $j$ vary with the biomass of this resource.
Consumers then lose biomass by being consumed, $e_ij$ being the assimilation
efficiency of the consumer $j$ on the resource $i$. Consumers also lose biomass
over time through metabolic losses ($-x_iB_i$) and by natural mortality due to
environment ($-d_iB_i$). The primary producers gain biomass over time with a
growth rate ($r_i$, eq. \@ref(eq:prod)) and a functional response ($G_i$)
describing how the growth of the producers vary with their biomass. They lose
biomass by being consumed by consumers and by natural mortality.

@@

The growth of the primary producers (eq. \@ref(eq:growth)) is logistic
where the growth rate is maximal and null when
$\sum_{j \in \{\mathrm{prod}\}} \alpha_{ij} B_j$ are respectively close to 0 and 1. $\alpha_{ij}$ is the
per capita effect of the producer $j$ on the producer $i$, $K_i$ being the
carrying capacity for the producer $j$. The functional response of a consumer
feeding on a resource (eq. \@ref(eq:holling)) depends on the relative preference
of the consumer on the resource ($\omega_ij$) is limited by a half-saturation
rate ($B_0$), intraspecific interference coefficient ($c_i$), and on the
availability of its other resources
($\sum_{k \in \{\mathrm{res}\}_i} \omega_{ik} B_k^h$). Finally, the $h$ exponent
controls the shape of the functional response from a saturating function
($h = 1$, Holling type II) to a sigmoid ($h = 2$, Holling type III).

@@

We introduced environmental stochasticity by considering stochastic variation of
mortality rates. Doing so allows to consider stochasticity without modifying
species interactions themselves
[@vasseur_environmental_2007]. The stochastic part of the mortality rates
($\epsilon_{i, t}$) follow a normal distribution 0 centered
($\epsilon_{i, t} \sim N(0,\sigma^2_e)$) with a variance ($\sigma^2_e$). The
mortality rates of each species at the time $t$ equals to $d_{i,t} = d_i
e^{\epsilon_{i, t}}$, which ensured that the mortality rates never become
negative [@vasseur_environmental_2007]. We
simulated $\epsilon$ using an Ornstein-Uhlenbeck process, which is a modified
Brownian motion where the stochastic values tend to come back to the central
value (i.e. $\epsilon = 0$) such as $\epsilon_t$ varies according to this
stochastic differential equation: $d\epsilon_t = (0 - \epsilon_t)d_t + \sigma_e
dW_t$, where $dW_t$ is a Brownian motion. The strength of environmental
stochasticity was then controlled by $\sigma_e$.

@@

\begin{equation}
\begin{aligned}
\epsilon_{i, t} =
  \begin{bmatrix}
     \rho_{11}\sigma^2_e & \rho_{12}\sigma^2_e & \cdots & \rho_{1n}\sigma^2_e \\
     \rho_{21}\sigma^2_e & \rho_{22}\sigma^2_e & \cdots & \rho_{2n}\sigma^2_e \\
     \vdots  & \vdots  & \ddots & \vdots  \\
     \rho_{n1}\sigma^2_e & \rho_{n2}\sigma^2_e & \cdots & \rho_{nn}\sigma^2_e
  \end{bmatrix}
  \begin{bmatrix}
     a_{1,t} \\
     a_{2,t} \\
     \vdots  \\
     a_{n,t}
  \end{bmatrix}
(\#eq:rho)
\end{aligned}
\end{equation}

@@

}
pS <- get_sim_par_val(x = "S")
pct <- round(get_sim_par_val(x = "ct"), 2)
pZ <- get_sim_par_val(x = "Z")
pc <- get_sim_par_val(x = "c")
pe <- get_sim_par_val(x = "env_stoch")
prd <- get_sim_par_val(x = "resp_div")
pr <- get_sim_par_val(x = "rho")
```
\pagebreak

@@

To vary interaction strength and mortality rates (i.e. allometrically
scaled mortality rate $d$) in the community, we varied the Predator-Prey
Mass Ratio (PPMR) from $`r pZ["min"]`$ to $`r pZ["max"]`$ (Fig. \@ref(fig:met)b).
We set $h = 2$, i.e. a functional response of type III (eq. \@ref(eq:holling)),
and varied predator interference (i.e. from $c = `r pc["min"]`$ to $c = `r
pc["max"]`$, eq. \@ref(eq:holling)), as predator interference was shown to have
a tremendous role on population stability [@brose_allometric_2006]. We set a global
carrying capacity of $K' = 10$ to ensure that consumers were not
limited by the biomass input from primary producers. We standardised the
carrying capacity by the number of primary producers and by the interspecific
competition among producers to ensure that the effects of species richness on
temporal stability are not driven by the increase in the number of primary
producers, thereby trivially increase biomass input for consumers. We then
standardised $K'$ such as $K_i = \frac{1 + \alpha_{ij} (S - 1)}{S}K'$
[@ives_stability_2000]. However, we found that our results were not affected
when removing the standardisation of carrying capacity (Fig. S1).

We numerically solved the stochastic differential equation systems during 2000
timesteps and collected the last 500 timesteps, similarly to previous studies
[@brose_allometric_2006]. We set $d_t = .1$, and $d_t = .05$ when instability
was detected during the simulations due to stiff changes in biomass. We set
extinction threshold to $10^{-6}$ and if an extinction happened in the last 500
timesteps, we continued to run the simulation for another 1000 timesteps until
there was no extinction events during the last 500 timesteps. When we found
disconnected species at the end of the simulation (i.e. a primary producer
without consumer or a consumer without prey), we set their biomass to 0 and
re-run the simulation for another 1000 timesteps. Except in the case of
disconnected species, we re-ran simulations with species starting biomass equals
to their biomass at the last timestep of the previous simulation. Supplementary
analysis showed that alternative choices of simulation processes, such as longer
simulations, rebuilding food-webs (i.e. resetting consumer preferences,
recomputing species bodymass according to the new food-web) after removing
disconnected species did not affect our results (Fig. S1). The food-web
generation and simulations were done using the `Julia` package
`EcologicalNetworkDynamics.jl` [@lajaaiti_ecologicalnetworksdynamicsjl_2024].

@@

We measured food-web properties that have been linked with stability in the
literature.  We measured total biomass, species richness, connectance, average
trophic level weighted by biomass, average omnivory and average interaction
strength. Food-web properties were measured at each of the 500 timesteps and
then averaged, except connectance and trophic level that were static.
Connectance was computed as $C = L / S(S-1)$, $L$ being the number of trophic
links and $S$ being the number of species in the community. Species trophic
level was computed recursively from the bottom of the food-web, such as the
trophic level of a consumer was equal to the average trophic level of its
resources added to 1, the trophic level of primary producers being set to 1
[@christensen_ecopath_1992]. Average trophic level was then equal to:
$1/\hat{B}_i\sum_{i}\mathrm{TL}_i$, $\mathrm{TL}_i$ being the trophic level and
$\hat{B}_i$ the average biomass of the species $i$. The degree of omnivory of a
consumer was computed  such as the sum of squares of its resource trophic levels
weighted by the relative preference of the consumer for each resource: $\sum_i
(\mathrm{TL}_i - \hat{\mathrm{TL}})^2 \omega_i/ \sum \omega_i$
[@christensen_ecopath_1992]. Interaction strength was computed was quantified as
the biomass fluxes going from a resource to a consumer, such as: $I = x_i y_i
B_k F_{ij}$ (eq. \@ref(eq:cons)), and averaged over time. Our way of measuring
interaction strength is very similar to previous metrics used in previous
studies, at the difference here that we average fluxes over time and not the
maximal interaction strength [@mccann_weak_1998]. Finally, the average
interaction strength of a community was computed as the average interaction
strength at the exclusion of null interaction strengths (i.e. excluding absence
of trophic interactions).

@@

To further get a mechanistic insight about the effects of stressors and food-web
structure on stability, we partitioned asynchrony into portfolio effects (measured by statistical averaging
effect, SAE) and compensatory dynamics effect (CPE) [@zhao_biodiversity_2022].
SAE quantities the portfolio effect emerging from independent species
fluctuations, while CPE describes compensatory dynamics deviating from portfolio
effects, and emerging from species interactions and differential responses to
environmental stochasticity.

@@

Using a structural equation model, we disentangled the complex effects of
food-web structure, environmental stochasticity and response diversity on the
different facets of temporal stability of biomass (i.e. population stability,
portfolio effects and compensatory dynamic effects). We tested the
effects of food-web structure (average interaction strength, average trophic
level, connectance) on population stability and asynchrony partitions (portfolio and
compensatory effects). Grounded in theoretical ecology, we expected that higher
average trophic level lower interaction strength, lower connectance[@mccann_weak_1998;
@mccann_reevaluating_1997; @shanafelt_stability_2018]. We further added
simulation parameters as control variables because higher Predator-Prey Mass
Ratio and predator interference were also expected to increase population
stability [@brose_allometric_2006]. Finally, we tested the effects of environmental
stochasticity and response diversity on population stability and asynchrony
partitions, although there were predicted to mainly affect to population stability
(i.e. by increasing amplitude of population fluctuations) and asynchrony (i.e.
by generating independent population fluctuations) respectively. We
completed the structural equation model with the mathematical relationship
linking compensatory dynamics and statistical averaging effects to asynchrony,
and from asynchrony and population stability to community stability. We further
developed a more detailled SEM including the effects omnivory and partition of
compensatory dynamics due to species interactions ($CPE_{int}$) and due to
environmental fluctuations ($CPE_{env}$).
